Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-27 2024/OPTIMA/OPTIMA_2024_MCO.tex
Optima part 3

\begin{equation}
\label{eq:04}
\min \varphi(y) = F( \lambda, y ), y \in D,
\end{equation}

\begin{equation}
\label{eq:05}
F(\lambda, y) = \max_{1 \leq i \leq s} {(\lambda_i f_i (y))}, y \in D, \sum_{i=1}^s {\lambda_i} = 1, 0 \leq \lambda_i \leq 1.
\end{equation}

\begin{equation}
\label{eq:06}
F(y) = \{ F(\lambda_1,y),F(\lambda_2,y), \dots ,F(\lambda_\tau,y)\}
\end{equation}

\begin{equation}
\label{eq:07}
\min{\varphi(y(x))} = \min {\varphi(y)}, x \in [0,1], y \in D
\end{equation}

\begin{equation}
\label{eq:08}
|\varphi(y(x')) - \varphi(y(x''))| \leq H \|x' - x''\|^{1/N} , x \in [0,1],
\end{equation}

>>>>

For solving the optimization problem (\ref{eq:02a}), in our study the computational scheme \cite{ML_MCO_2023} has been applied. It consists of the following main stages.

\begin{enumerate}
	\item Scalarization of the vector criterion (\ref{eq:01}).
	\item Dimensionality reduction.
	\item Optimization of transformed functions.
\end{enumerate}
Below is a brief description of each stage.

\textbf{Scalarization.} It is possible to find the Pareto set of optimal parameters either by solving the problem directly in the initial statement \cite{Evtushenko2014,Deb2002,Durillo2010,Mostaghim2007,NDG09,RC05,ZLT01} or by transforming it to a series of scalar subproblems when each subproblem provides finding separate efficient point \cite{Pardalos2017,ML_MCO_2023,Gergel2019_2,Gergel2018,GergelKozinov2020}.

Within the framework of the described research, the second approach is used. In general, the optimization problem generated by scalarization of the vector criterion (\ref{eq:01}) can be written in the form

\begin{equation}
\label{eq:04}
\min \varphi(y) = F( \lambda, y ), y \in D,
\end{equation}
where $F$ is a scalar function including partial criteria $f_i$, $1 \leq i \leq s$, and the vector
\begin{equation}
\label{eq:04a}
\lambda = (\lambda_1, \lambda_2, \dots, \lambda_s), \lambda_i > 0, 1 \leq i \leq s, \sum_{i = 1}^s {\lambda_i} = 1,
\end{equation}
reflects the ``importance'' (weights) of corresponding partial criteria.

There are various ways to construct the function $F$ \cite{Miettinen1999,Ehrgott2005,GergelKozinov2020,Marler2004}. In our study we utilized the maximum convolution 
\begin{equation}
\label{eq:05}
F(\lambda, y) = \max_{1 \leq i \leq s} {(\lambda_i f_i (y))}, y \in D, \sum_{i=1}^s {\lambda_i} = 1, 0 \leq \lambda_i \leq 1.
\end{equation}

%\begin{equation}
%\label{eq:06}
%F(y) = \{ F(\lambda_1,y),F(\lambda_2,y), \dots ,F(\lambda_\tau,y)\}
%\end{equation}

If all the partial criteria are Lipschitzian and at least one function $f_i(y)$, $1 \leq i \leq s$, is multiextremal, the convolution (\ref{eq:05}) will be Lipschitzian and multiextremal as well.

In the case of partial criteria positivity the global minimizer of the convolution (\ref{eq:05}) is the Pareto-optimal point. In order to obtain all elements of the Pareto set, it is necessary to solve the series of scalar problems (\ref{eq:04}) for all weight vectors $\lambda$ satisfying the conditions from (\ref{eq:05}).

\textbf{Dimensionality reduction.} For solving the optimization problems (\ref{eq:05}) the dimensionality reduction \cite{Gergel2019_2,Gergel2018,GergelKozinov2020} on the base of Peano curves is applied. The Peano curve $y(x)$ maps the close interval $[0,1]$ of the real axis onto the parallelepiped (\ref{eq:02}) unambiguously and continuously. It enables to replace the multidimensional problem (\ref{eq:04}) with solving the equivalent univariate problem
\begin{equation}
\label{eq:07}
\min{\varphi(y(x))} = \min {\varphi(y)}, x \in [0,1], y \in D
\end{equation}

It should be noted that if the convolution $F(\lambda,y)$ satisfies the Lipschitz condition then the reduced function (\ref{eq:02}) meets the H{\" o}lder condition

\begin{equation}
\label{eq:08}
|\varphi(y(x')) - \varphi(y(x''))| \leq H \|x' - x''\|^{1/N} , x \in [0,1],
\end{equation}
with the H{\" o}lder constant $H=2L\sqrt{N+3}$, where $L$ is the Lipschitz constant of $F(\lambda,y)$ in the domain $D$.
